plyranges: a grammar of genomic data transformation

### 2018-10-24 paper.Rmd
declarations and software versioning

---
title: "plyranges: a grammar of genomic data transformation" 
author:
  - name: Stuart Lee \footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}
    email: stuart.lee1@monash.edu
  - name: Dianne Cook $^{\ref{monash}}$
    email: di.cook@monash.edu 
  - name: Michael Lawrence \footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author} 
    email: lawrence.michael@gene.com
abstract: "There is a cognitive load placed on users in learning a data abstraction from the Bioconductor project and understanding its appropriate use. Users must navigate these abstractions to perform a genomic analysis task, when a single data abstraction, a GRanges object will suffice. By recognizing that the GRanges class follows 'tidy' data principles, we create a grammar of genomic data transformation, defining verbs for performing actions on and between genomic interval data and providing a way of performing common data analysis tasks through a coherent interface to existing Bioconductor infrastructure. We implement this grammar as a Bioconductor/R package called \texttt{plyranges}."
keywords: "Bioconductor, Grammar, Genomes, Data Analysis" 
bibliography: references.bib 
output: 
  pdf_document:
    template: svm-latex-ms.tex 
    keep_tex: true 
    fig_caption: true
    latex_engine: pdflatex 
    citation_package: biblatex 
fontfamily: mathpazo
---

```{r setup, include = FALSE}
library(tibble)
library(knitr)
opts_chunk$set(message = FALSE, 
               warning = FALSE, 
               comment = "#>", 
               fig.width = 5, 
               fig.height = 3, 
               fig.align = "center")
# for peak figures
library(ggbio)

@@

# Methods
## Data Availability
The BigWig file for the H3K27Me3 primary T CD8+ memory cells from peripheral 
blood ChIP-seq data was downloaded from the `AnnotationHub` package (2.12.0)
under accession AH33458. The BAM file corresponding to the
H1 cell line ChIP-seq data is available at GEO under accession GSM433167. 
The SNP array data for the H1 cell line data is available at GEO under accession
GSM1463263.

## Software Versions
To produce the workflows as described in results section we used 
R version 3.5 with the development version of `plyranges` (1.1.10) 
and the  `BioStrings` package (2.48.0) installed.

@@

# Acknowledgements

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages
for the code review they performed. This article was written with
`knitr` \cite{R-knitr} and the figures were made with `ggbio` \cite{R-ggbio}.

>>>>

---
title: "plyranges: a grammar of genomic data transformation" 
author:
  - name: Stuart Lee \footnote{\label{monash}Department of Econometrics and Business Statistics, Monash University, Clayton, Australia}\footnote{Molecular Medicine Division, Walter and Eliza Hall Institute, Parkville, Australia}
    email: stuart.lee1@monash.edu
  - name: Dianne Cook $^{\ref{monash}}$
    email: di.cook@monash.edu 
  - name: Michael Lawrence \footnote{Bioinformatics and Computational Biology, Genentech Research and Early Development, South San Francisco, United States of America}\footnote{corresponding author \href{mailto:michael.lawerence@gene.com}{michael.lawerence@gene.com} } 
    email: lawrence.michael@gene.com
abstract: "There is a cognitive load placed on users in learning a data abstraction from the Bioconductor project and understanding its appropriate use. Users must navigate these abstractions to perform a genomic analysis task, when a single data abstraction, a GRanges object will suffice. By recognizing that the GRanges class follows 'tidy' data principles, we create a grammar of genomic data transformation, defining verbs for performing actions on and between genomic interval data and providing a way of performing common data analysis tasks through a coherent interface to existing Bioconductor infrastructure. We implement this grammar as a Bioconductor/R package called \texttt{plyranges}."
keywords: "Bioconductor, Grammar, Genomes, Data Analysis" 
bibliography: references.bib 
output: 
  pdf_document:
    template: svm-latex-ms.tex 
    keep_tex: true 
    fig_caption: true
    latex_engine: pdflatex 
    citation_package: biblatex 
fontfamily: mathpazo
---

```{r setup, include = FALSE}
library(tibble)
library(knitr)
opts_chunk$set(message = FALSE, 
               warning = FALSE, 
               comment = "#>", 
               fig.width = 5, 
               fig.height = 3, 
               fig.align = "center",
               fig.path = "diagrams")
# for peak figures
library(ggbio)

@@

# Methods
## Data Availability

The BigWig file for the H3K27Me3 primary T CD8+ memory cells from peripheral 
blood ChIP-seq data from the Human Roadmap Epigenomics project  
was downloaded from the `AnnotationHub` package (2.13.1)
under accession AH33458 \cite{R-ahub, Roadmap-Epigenomics-Consortium2015-pr}. 
The BAM file corresponding to the H1 cell line ChIP-seq data is available at 
NCBI GEO under accession [GSM433167](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM433167) \cite{ncbi-geo, Roadmap-Epigenomics-Consortium2015-pr}. 
The SNP array data for the H1 cell line data is available at  NCBI GEO under accession
[GPL18952](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GPL18952) 
\cite{Roadmap-Epigenomics-Consortium2015-pr}.

## Software Versions
To produce the workflows as described in results section we used 
R version 3.5 with the development version of `plyranges` (1.1.10)   
and the `BioStrings` package (2.49.0) installed \cite{r-core, plyranges-zenodo, R-biostrings}.

@@

# Declarations

## Acknowledgements

We would like to thank Dr Matthew Ritchie at the Walter and Eliza Hall 
Institute and Dr Paul Harrison for their feedback on earlier drafts of
this work. We would also like to thank Lori Shepherd and Hèrve Pages
for the code review they performed and users who have submitted feedback
and pull requests. This article was written with `knitr` \cite{R-knitr} and 
the figures were made with `ggbio` \cite{R-ggbio}.

## Funding

SL is funded by an Australian Government Research Training Program (RTP) 
scholarship and receives an additional top up scholarship from CSL. Funding 
bodies had no role in data collection, analysis, and interpretation.

## Authors Contributions

SL, DC, ML contributed ideas towards the software. SL wrote 
the manuscript and analysed data. SL and ML wrote the software. 
DC and ML supervised the research and revised the manuscript. 
All authors have read and approved the final version of the manuscript.

## Ethics approval and consent to participate

Not applicable to the current manuscript.

## Competing Interests

Not applicable to the current manuscript.
